---
title: "Chapter 5_Working_With_Numbers"
author: "Gustavo R Santos"
date: '2022-07-11'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 5 - Working with Numbers

This document is part of the Packt Book *Data Wrangling with R*.

---

**Creating Numeric variables**
```{r}

# Creating a scalar
scalar <- 42
print(scalar)

# Creating a vector
vec <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
print(vec)

# Creating a Matrix
mtrx <- matrix(data=vec, nrow = 3, ncol= 3)
print(mtrx)

# Creating Data frame
df <- data.frame(column1= c(1,2,3),
                 column2= c(4,5,6),
                 column3= c(7,8,9))
print(df)

```


**Vectors**
```{r}

# Vector
vec1 <- c(9,8,7,6,5,4,3,2,1)
vec2 <- c(1,2,3,4,5,6,7,8,9)

# Slicing a vector
vec1[1:3]

# Slice by condition
vec1[vec1 >= 6]

# Sum of vectors
vec1 + vec2

# Multiplication of vectors
vec1 * vec2

# Addition or Multiplication of a vector by a single number
vec1 * 10
vec1 - 1

# Two vectors of different sizes
c(1,2) + c(1,2, 3, 4)

```

**Matrix**
Two dimensional elements that require a different slicing notation

```{r}

# Create a Matrix
mtrx <- matrix(1:12, nrow=4, ncol=3, byrow=T)

# Slicing [row,col]
mtrx[2,3]


```
**Data Frames**
They are heterogeneous. They accept variables with different data types.
```{r}

# Data frame
df <- data.frame(name = c('Carl', 'Vanessa', 'Hanna', 'Barbara'),
                 class_ = c('Math', 'Math', 'Math', 'Math'),
                 grade = c(8.5, 9, 9, 7)   )


# Slicing Carl's grade (col 1 and col 3)
df[1,3]

# Slicing by variable
df[, "grade" ]

# Slicing by condition
df[ df$grade > 8,]

```

Example: Extra point from the "professor"
```{r}

# Extra point
# Scenario: everyone delivered
df$new_grade <- df$grade + 1

df
```
Normalization example

```{r}

# Normalization
max_grade <- max(df$grade)
df$grade <- df$grade/max_grade

df
```

**Apply Functions**

`apply()`
 It let’s you apply a function to a row or a column of a data frame. 
 Use margin = 1 to apply the function to every element of a row and 2 to apply it to every element of a column.
 
```{r}

# apply function mean to rows: sum 'grade' + 'new_grade'

apply(df[,c('grade', 'new_grade')], MARGIN= 1, FUN = mean)

# apply function mean to columns the column 3 and 4
apply(df[,c(3,4)], MARGIN= 2, FUN= mean)

# apply custom function
my_func <- function (x){sum(x)/2}
apply(df[,c(3,4)], MARGIN= 1, FUN= my_func)

```
 
 `lapply()`: List apply
Used to apply a function to lists of objects, returning a list object of the same length.

```{r}

# lapply of a custom function to remove the extra point from new_grade
my_func <- function (x){x-1}
lapply_obj <- lapply(df$new_grade, my_func)

```


 `sapply()`: Same Length Apply
Used to apply a function to lists of objects, returning an array or matrix object of the same length.

```{r}

# sapply of a custom function to a single column
my_func <- function (x){x-1}
sapply(df$new_grade, my_func)

# sapply of a custom function to two columns
sapply(df[,c(3,4)], my_func)

```

 `tapply()` : Text apply
 This one applies your function to each factor variable of a vector.
 
```{r}

# Create new df
# Data frame
df <- data.frame(name = c('Carl', 'Vanessa', 'Hanna', 'Barbara', 'Jason', 'Alison', 'Kevin', 'Melody'),
                 class_ = c('Math', 'Math', 'Math', 'Math', 'Math2', 'Math2', 'Math2', 'Math2'),
                 grade = c(8.5, 9, 9, 7, 5, 7, 10, 9.5)   )

# Calculate the mean for each class_
tapply(df$grade, df$class_, mean)

```
 
 ---
 
 ## Descriptive Statistics
 
 Descriptive statistics are measurements or summarizations of a sample used to describe a variable
```{r}

# Set a seed to reproduce the same results
set.seed(12)

# Creating a variable: 50 data points, mean 0, standard deviation 1
var1 <- rnorm(50, mean = 0, sd=1)

# Creating a variable: 50 data points, mean 0, standard deviation 10
var2 <- rnorm(50, mean = 0, sd=5)

# Mean
mean(var1)
mean(var2)

# Standard Deviation
sd(var1)
sd(var2)

# Median
median(var1)
median(var2)

# percentiles or quantiles
quantiles(var1, c(0.25, 0.5, 0.75))




```
 
 
 # Histograms
 
```{r}


# Setup grid with 2 columns
par(mfrow=c(1,2))

# Histogram var1
hist(var1, col='royalblue', ylim=c(0,20), xlim=c(-20,20) )

# Histogram var1
hist(var2, col='coral', ylim=c(0,20), xlim=c(-20,20) )

# Reset grid to single plot
par(mfrow=c(1,1))

```
 
Descriptive statistics in numbers
```{r}

# Create a dataframe with both variables for comparison
df <- data.frame(var1= var1,
                 var2= var2)



# Comparison data frame
data.frame( avg= apply(df, 2, mean),
            std_dev= apply(df, 2, sd),
            min_val= apply(df, 2, min),
            pct25= apply(df, 2, function (x) {quantile(x,0.25)}),
            median_val= apply(df, 2, median),
            pct75= apply(df, 2, function (x) {quantile(x,0.75)}),
            max_val= apply(df, 2, max)
            )



```

Calculate statistics using the function `summary()`.

```{r}

# Descriptive stats with summary function
summary(df)

```


```{r}
library(skimr)
skim(df)
```

**Correlation**
Correlation is the measurement of the strength of a linear relationship between two variables.
```{r}

# Correlation
cor(var1, var2, method='pearson')

# Scatterplot
plot(var1, var2, col='royalblue', pch=16,
     main='Scatterplot var1 x var2')

```




