---
title: "Chapter7_Transformations_Base_R"
author: "Gustavo R Santos"
date: '2022-07-21'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 7 - Transformations with base R

This document is part of the Packt Book *Data Wrangling with R*.

---

## Dataset

The dataset to be used in this chapter is the *Adult Data Set* from the UCI Machine Learning Repository.

**Dataset Credits:**
Dua, D. and Graff, C. (2019). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science.

**URL Address**
https://archive.ics.uci.edu/ml/datasets/Adult


### Import Libraries
```{r}
library(data.table)
library(stringr)
```


### Loading the dataset

```{r}

# Load the dataset to RStudio
df <- read.csv('https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data',
               header=FALSE, strip.white = TRUE)

# Add column names
colnames(df) <- c('age', 'workclass', 'fnlwgt','education', 'education_num', 
                  'marital_status', 'occupation', 'relationship', 'race', 'sex',
                  'capital_gain', 'capital_loss','hours_per_week', 'native_country', 'target')

head(df)

```


#### Slicing

Slicing is like zooming in our dataset, looking only the parts we need or want to see.

```{r}

# Dataset dimensions
dim(df)

# Slicing rows 1 to 3, all columns
df[1:3,]

# Slicing all rows, columns 1 to 3
df[,1:3]

# Slicing with conditional, vector for cols 1 and 6
df[df$age > 30 ,c(1,6)]

```

#### Filtering

Filter differs from slice because the filter returns all the variables, so it won't be able to cut out some variables.

```{r}

# Filter
subset(df, age > 30)


# Slice with subset, returning only the selected columns
subset( df, age > 30, select=c(age, marital_status) )

```

### Group by and summarize
Group by will put the data in groups. Summarize will reduce the aggregated data values to one number, like mean, median, count etc.

```{r}

# Data frame to Data.table
dt <- as.data.table(df)

# Group By and count number of observations
dt[, .N, by = sex]

# Group By and return mean of age by sex
dt[, mean(age), by = sex]

# Group By and return sum of capital_gain by sex
dt[, sum(capital_gain), by = sex]


```

**Group and visualize**
```{r}

# Mean education years by education level
educ_yrs <- dt[, mean(education_num), by= education]
educ_yrs <- educ_yrs[order(V1)]

# barplot
barplot(educ_yrs$V1, names.arg = educ_yrs$education, col='royalblue',
        main='Average years of education by grade level')


```

### Replace and Fill values

#### Replace

Replace is to exchange a value for another.
Fill values is when you have a `NA` value that needs to be filled with an average or another value, like zero.

```{r}

# Check for NA values
sum(is.na(df))
```

```{r}
# Replace "?" using slicing
df$occupation[df$occupation == '?'] <- NA
df[df$native_country == '?', 'native_country'] <- NA

# Replace "?" with NA using gsub
df$workclass <- gsub(pattern= '\\?', replacement= NA, df$workclass)

# Replace "?" with ifelse
df$native_country <- ifelse(df$native_country == '?', NA, df$native_country)

# Check for NA values
sum(is.na(df))

```
**Check missing data by column**
```{r}

library(skimr)

skim(df)

```

#### Filling
Common fill values are zero, mean, median, most frequent (for categorical).
There are more complex ways to fill, using machine learning, but that is not in the scope of this book.

```{r}

#Fill NAs with most frequent value
most_frequent <- names(table(df$workclass)[which.max(table(df$workclass))])
df$workclass[is.na(df$workclass)] <- most_frequent

# Occupation
most_frequent <- names(table(df$occupation)[which.max(table(df$occupation))])
df$occupation[is.na(df$occupation)] <- most_frequent

# Remove all NAs (the remaining in native_country)
df <-  na.omit(df)

```

**Filling NAs from Numeric columns**

```{r}

# Data frame
df_num <- data.frame(A= c(1,2,2,2,3,NA),
                     B= c(3,4,5,3,NA,0),
                     C= c(1,1,1,NA,NA,5))

# Fill NAs with 0
nafill(df_num, type='const', fill=0)

# Fill NAs with the last valid number
nafill(df_num, type='locf')

# Fill NAs with the next valid number
nafill(df_num, type='nocb')

# Fill NAs with mean
df_num$A[is.na(df_num$A)] <- mean(df_num$A, na.rm = T)

```


### Arranging Data
Arrange data is useful to rank items from top to bottom or vice-versa, creating ordinated datasets.

```{r}

# Arrange raw data
df[order(df$age, decreasing = T),]

# Arrange raw data using - instead of "decreasing=T"
df[order(-df$age),]


```
Using data.table, we can group variables and order them.

```{r}

# Data frame to Data.table
dt <- as.data.table(df)

# Group and order education years average by workclass
dt[, mean(education_num), by= workclass][order(-V1),]


```


### Creating New Variables
During data wrangling, it is common to create new variables, being that for splitting an information or combining two or more variables to make a new one.

Note: to create new columns in R, you can use `df$new_col_name` or `df['new_col_name]`

```{r}

# Binarizing sex using ifelse()
df$sex_binary <- ifelse(df$sex == 'Female', 1, 0)
df[,c(10,16)]

```

To Encode a variable, transforming the names in numbers, here is a solution.
We will: 
1. Assign the variable `education` as a factor
2. Create order to the levels
3. Store those orders in another variable
4. Change the new variable to numbers

```{r}
## Encoding a variable ##
# Assign education as factor
df$education <- as.factor(df$education)
# Order the factors
df$education <- ordered(df$education, levels= c("Preschool", "1st-4th","5th-6th","7th-8th", "9th", "10th", "11th", "12th", "HS-grad", "Some-college", "Assoc-acdm", "Assoc-voc", "Bachelors", "Masters" , "Doctorate") )
# Create new variable
df$education_cd <- df$education
# Change levels to numbers
levels(df$education_cd) <- 1:15

df[, c('education', 'education_cd')]

```

**Creating a custom calculation**
Imagine that there were a rule in place that, for every `capital_gain` - `capital_loss` equal or greater than 15,000 there was a 10% tax.
We will first create a `total` variable to show arithmetic operation and then calculate the 10% tax

```{r}

# Total gain variable creation
df$total_gain <- df$capital_gain - df$capital_loss

# Tax variable creation
df$tax <- ifelse(df$total_gain >= 15000, df$total_gain *0.1, 0)

df[,c(11,12,17,18)][order(-df$tax),]

#---
#Shorter code

# Tax variable creation
df$tax2 <- ifelse(df$total_gain >= 15000, (df$capital_gain - df$capital_loss) *0.1, 0)


```

### Binding

Binding data is useful to glue pieces of data together.

`rbind()` or bind rows glues the data by columns. So if your data has the same variables, you can use this function.
```{r}

# Creating datasets A and B
A <- df[1:10, ]
B <- df[11:20, ]

# Append / bind rows
AB <- rbind(A, B)
AB

```



`cbind()` is to bind columns, gathering the data by rows. It is more suitable when the rows are for the same observation, but the columns are different and need to be put together.

```{r}

# Creating datasets A and B
A <- df[1:10, 1:3]
B <- df[1:10, c(4,7)]

# Bind columns
AB <- cbind(A, B)
AB

```

### data.table
Review of the library `data.table`

Basic syntax
DT[i, j, by]
* *i* is for the rows selection or a condition for the rows to be displayed
* *j* is to select the variables or to calculate a statistic of it
* *by* is used to group variables

```{r}

# Assign variable as data.table object
dt <- as.data.table(df)

# Syntax
dt[dt$age > 50, .(age, occupation)]

# Mean age Group By relationship
dt[, mean(age), by= relationship][order(-V1),]


```


