---
title: "Chapter4_Working_Regex"
author: "Gustavo R Santos"
date: '2022-07-04'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 2 - How to Load Files to RStudio

This document is part of the Packt Book *Data Wrangling with R*.

---

# StringR
This library is part of the `tidyverse` package and brings you many tools to work with strings.

### **Import Libraries**
```{r}

# Use install.packages('library_name') is you don’t have it installed
library(tidyverse)
library(stringr)
library(gutenbergr)

```

### **Escape Patterns**

We can use an escape character – double backslashes \\ – to tell the computer: “hey, if I say \\., I am actually trying to look for a dot”. You can run this code `?"'"` to see other escape patterns.

```{r}

# Run this code to look for escape patterns
#?"'"

# text
txt <- 'Looking for a .'

# Regex escape
str_view(txt, pattern='\\.')
```

Find a single character.

```{r}

# text
txt <- 'This is a text for us to learn RegEx 101! Ok?'

# Find a single character
str_view_all(txt, '[r]')

```

Find numbers, text or punctuation

```{r}

# Find numbers
str_view_all(txt, '[:digit:]')

# Find letters
str_view_all(txt, '[:alpha:]')

# Find punctuation
str_view_all(txt, '[:punct:]')

# Find spaces
str_view_all(txt, '[:space:]')

```

### **Basic codes**

```{r}

# text
txt <- 'This is a text for us to learn RegEx 101! Ok?'

# Find a single character
str_view_all(txt, '[a-z]')

# Not match is
str_view_all(txt, '[^is]')

# Find a digits
str_extract_all(txt, '\\d')

# Find not digits
str_extract_all(txt, '\\D')

# Find words
str_extract_all(txt, '\\w')

# Find not words
str_extract_all(txt, '\\W')

# Find white spaces
str_extract_all(txt, '\\s')

# Find not white spaces
str_extract_all(txt, '\\S')

# Begins with T
str_view_all(txt, '^T')

# Ends with k
str_view_all(txt, '\\?$')

# Boundaries
str_view_all(txt, '\\btext\\b')

# Pattern can happen zero or more times
str_view_all(txt, '[0*]')

# Pattern can happen one or more times
str_view_all(txt, '1+')

```




Let's download this public domain book Alice’s Adventures in Wonderland, by Lewis Carrol.
We will use the text to learn more about stringr.

**Download from Gutenberg library**
```{r}

# Download Alice's Adventures in Wonderland
#gutenberg_metadata
alice <- gutenberg_download(gutenberg_id=11)

```

If we want to know how many times the word Alice happens in the text, here is a possible solution.

```{r}

# Exact Match of Alice
sum(str_count(alice$text, 'Alice'))

# How many digits in the text?
sum(str_count(alice$text, '[:digit:]'))

# How many written numbers 1-5. | means “or”.
sum(str_count(alice$text, 'one|two|three|four|five'))

# How many words in UPPERCASE
sum(str_count(alice$text, '\\b[A-Z]+\\b'))

# Count the gerund words, ending in "ing"
sum(str_count(alice$text, '\\b(\\w+ing)'))


```

List the gerund words
```{r}

# Extract the gerund words, ending in "ing"
gerunds <- c( str_extract_all(alice$text, '\\b(\\w+ing)') )

# Show only the values
unlist(gerunds)

```



