---
title: "Chapter4_Data_Summaries"
author: "Gustavo R Santos"
date: '2022-07-06'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 2 - How to Load Files to RStudio

This document is part of the Packt Book *Data Wrangling with R*.

---


Let's download this public domain book Alice’s Adventures in Wonderland, by Lewis Carrol.
We will use the text to learn more about stringr.

**Download from Gutenberg library**
```{r}

# Download Alice's Adventures in Wonderland
alice <- gutenberg_download(gutenberg_id=11)

```

Creating a data summary of the words ending with *"ing"* in the book

```{r}

# Extract the "ing" ending words
ings <- str_extract_all(alice, '\\b(\\w+ing)')

# Transform the words in a vector
ings <- unlist(ings)

# Data Frequency to data.frame
df_ings <- data.frame( table(ings) )

# Sorting descending and collecting the top 10 most frequent observations
df_ings_top10 <- df_ings[order(-df_ings$Freq)[1:10],]

```

Creating a complete data summary of the main characters of the book

```{r}

# List characters
characters <- c('Alice', 'Rabbit', 'Queen', 'King', 'Cheshire Cat', 'Duchess', 'Caterpillar', 'Hatter')

# Create regex string
char_regex <- str_c(characters, collapse = '|')

# Make the entire text Title Case for better match
alice_title_case <- str_to_title(alice$text)

# Extract the words
count_book_chars <- unlist( str_extract_all(alice_title_case, char_regex) )

# Data Frequency to data.frame
df_book_chars <- data.frame( table(count_book_chars) )

```

Enhance the frequency table

```{r}

# Sort table by frequency
df_book_chars <- df_book_chars[order(-df_book_chars$Freq),]

# Add percentages rounded to 3 decimals
df_book_chars$pct <- round( proportions(df_book_chars$Freq), 3)

# Add cumulative sum of the pct
df_book_chars$pct_cumSum <- cumsum(df_book_chars$pct)

```


```{r}

# Install
#install.packages("gmodels")
# Load gmodels
library(gmodels)

#Load the data
data("mtcars")

# Create the contingency table
CrossTable(mtcars$cyl, mtcars$gear, prop.t=TRUE, prop.r=TRUE, prop.c=TRUE)

```


### Factors

Factor is a type of object to create categorical variables in R.
We can create factors with ordered levels or not.

```{r}

# Textual variable
var <- c('A', 'B', 'B', 'C', 'A', 'C')

# To create a factor
factor_var <- factor(var)

```

To create levels with hierarchical order, use the following code `levels()`.
The order created was C < B < A
```{r}

# Ordered levels
levels(factor_var) <- c('C','B','A')

factor_var

```



