---
title: "Chapter4_Working_With_Strings"
author: "Gustavo R Santos"
date: '2022-07-03'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 2 - How to Load Files to RStudio

This document is part of the Packt Book *Data Wrangling with R*.

---

# StringR
This library is part of the `tidyverse` package and brings you many tools to work with strings.

### **Import Libraries**
```{r}

# Use install.packages('library_name') is you don’t have it installed
library(tidyverse)
library(stringr)
library(gutenbergr)

```

### **Creating a string**
To create a string, use single or double quotes 
```{r}

string1 <- 'I am a string'
string2 <- "me too"
string3 <- 'quote inside "quote", use single quote for the string and double for the text.'

# Priting
print(string1)
writeLines(string2)
writeLines(string3)

```

### **Detecting Patterns**
Detect if a string has a given pattern.

```{r}

# Create a string
text <- 'hello'
# Detect if the string has the letters "rt"
str_detect(text, 'rt')
# Detect if the string has the letters "ll"
str_detect(text, 'll')

# Create a phrase
text <- 'hello, world!'
# Detect if the string has the letters "ll"
str_detect(text, 'world')


```

Determine if a string starts with a pattern.
This can be useful to classify texts, for example

```{r}

# Create strings
my_id <- 'ORDER-1234'

# Starts with
str_starts(string=my_id, pattern='ORDER')

```
Find the index of a pattern

```{r}

# Create vector
shop_list <- c('fruit', 'vegetable', 'pasta')

# Index of "pasta"
str_which(string=shop_list, pattern='pasta')

```
Find location of a pattern

```{r}

shop_list <- c('fruit', 'vegetable', 'pasta')

str_locate(shop_list, 'pasta')

```
Counting pattern

```{r}

# Text
text <- 'I want want want to count the repetitions of want in this phrase. Do you want the same?'
# Count occurrences of want
str_count(text, 'want')

```

### **Subset Strings**

Extract only the text between the start and end defined in the function.
```{r}
# Create strings
my_id <- 'ORDER-1234'
# Count occurrences of want
str_sub(my_id, start=1, end=5)
```

Return only the matches

```{r}
# Text
my_ids <- c('ORDER-1234', 'ORDER-2234', 'MAINT-1234', 'MAINT-2234')

# Return orders
str_subset(my_ids, 'ORDER')

```


### **Manage Lengths**

Length of Strings

```{r}

# text
text <- 'What is the size of this string?'
# Length
str_length(text)

```
Trim

```{r}
# text
text <- ' Text    to  be trimmed. '

#trim
str_trim(text, side='both')

# squish
str_squish(text)

```

### **Mutate Strings**

Change letter case

```{r}
# text
text <- 'Hello world.'

# to UPPERCASE
str_to_upper(text)

# to lowercase
str_to_lower(text)

# to Title Case
str_to_title(text)

```

Replace a pattern

```{r}
# text
text <- 'Hello world. The world is beautiful!'

# Replace a pattern
str_replace(text, 'world', 'day')

# Replace all the patterns at once
str_replace_all(text, 'world', 'day')

```
### **Join and Split**

Join
```{r}

# text
s1 <- 'Hello'
s2 <- 'world!'

# concatenate
str_c(s1, s2, sep=' ')

```

Split string

```{r}

#text
text <- 'I am learning how to split strings'

# split
str_split(text, pattern=' ')

```

Order strings
```{r}

# text
shop_list <- c('bananas', 'strawberries', 'avocado', 'pasta')

# ordinate
str_sort(shop_list, decreasing = FALSE)

```

