---
title: "Chapter8_Transformations_tidyverse"
author: "Gustavo R Santos"
date: '2022-07-26'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 8 - Transformations with `tidyverse`

This document is part of the Packt Book *Data Wrangling with R*.

---

## Dataset

The dataset to be used in this chapter is the *Adult Data Set* (also know as *Census Income*) from the UCI Machine Learning Repository.

**Dataset Credits:**
Dua, D. and Graff, C. (2019). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science.

**URL Address**
https://archive.ics.uci.edu/ml/datasets/Adult


### Import Libraries
```{r}

library(tidyverse)
```


### Loading the dataset

```{r}

# Define column names
header <- c('age', 'workclass', 'fnlwgt','education', 'education_num', 
               'marital_status', 'occupation', 'relationship', 'race', 'sex',
               'capital_gain', 'capital_loss','hours_per_week',
               'native_country', 'target')

# Load the dataset to RStudio
df <- read_csv('https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data', col_names = header , trim_ws = TRUE)

head(df)

```


#### Slicing

Slicing is like zooming in our dataset, looking only the parts we need or want to see.

```{r}

# Slicing rows 1 to 5, columns 1 to 4.
df %>% .[1:5, c(1:4)]

# Slicing with slice_min() and slice_max()
df %>% slice_min(age, prop=0.10)
df %>% slice_max(age, prop=0.30)

# Slice sample
df %>% slice_sample(n=10, replace=TRUE)

```

#### Filtering
Filter returns all the variables for the observations that fulfill the filter condition. Using `select()`, we can also select only the wanted variables.

```{r}

# Filtering age over 30 years old
df %>% filter(age > 30)

# Filter age > 30 and selecting age and marital_status
df %>% 
  filter(age >30) %>% 
  select(marital_status, age)

# Distinct - removing duplicates
df %>% distinct(sex)

```

### Group by and summarize
Group by will put the data in groups. Summarize will reduce the aggregated data values to one number, like mean, median, count etc.

This is how the functions behave when called alone
```{r}

# group by not summarized
df %>% group_by(workclass)

# summarise without group_by returning average age of the dataset
df %>% summarise( age_avg = mean(age) )

# Group By workclass and Summarize mean age
df %>% 
  group_by(workclass) %>% 
  summarise(age_avg = mean(age))


```

Let's look at the object returned by dataset grouped by more than one variable. Notice that it is a grouped tibble. To make the return object be a regular tibble, use the function `ungroup()`.
```{r}

# Returns object grouped_df
workclass_groups <- df %>%
  group_by(workclass, sex) %>% 
  summarise(age_avg = mean(age) )

# Returns object tibble
workclass_ungrouped <- df %>%
  group_by(workclass, sex) %>% 
  summarise(age_avg = mean(age) ) %>% 
  ungroup()


```
**Summary Functions**
Functions to be used with `summarise()`.

```{r}

# n() shows the count of observations in each group
df %>% group_by(workclass) %>% summarise(n())

# n_distinct() shows the count of unique observations in each group
df %>% group_by(workclass) %>% summarise(n_distinct())

# sum(!is.na()) shows the count of Non NA observations
df %>% summarise(sum(!is.na(workclass)))

# first() shows the first age value in each group
# Similarly, you can use last() or nth()
df %>% group_by(workclass) %>% summarise(first(age))

# quantile() shows the top number that meets the quantile percentage chosen
# In this case, 50% of the age observations are under what value by group
df %>% group_by(workclass) %>% summarise(quantile(age, 0.5))

# sd() shows standard deviation of a variable
df %>% group_by(workclass) %>% summarise(sd(capital_gain))

# Across function, to apply the function mean to the selected columns
df %>% select(1,3,5,11,12,13) %>% 
  summarise(across(everything(), mean))

```

### Replace and Fill values
For this part of the exercise, let's create a custom function to count the NAs by column. It will be useful during the exercise.

```{r}

na_by_column <- function(df){
  "Function that takes in a data frame and return a table with the NAs count by column.
  * Input:
  df: data frame
  * Return
  data.frame object
  "
  # New data frame to hold the results
  df_na <- data.frame()
  # Loop through variables, count NAs and add to a row
  for ( variable in 1: length(colnames(df)) ){
    df_na[variable,1] <- sum( is.na(df_replaced[variable]) )
    }#close for loop
  # change row names
  rownames(df_na) <- colnames(df)
  # change column name
  colnames(df_na) <- 'NA count'
  # return result
  return(df_na)
  
}#close function

```

#### Replace

Replace is to exchange a value for another. Here, we will replace the `?` with `NA`, so we can fill the NAs later.
The first way to do that is using replace. 

```{r}

# Loop through variables looking for cells == "?"
for (variable in colnames(df)){
  print(
    paste( variable,
           dim( df[df[variable]=='?', variable])[1] )
  )
}

# Replacing values "?" with NA, saving in a new dataset variable
df_replaced <- df %>% 
  mutate(workclass = replace(workclass, workclass == '?', NA),
         occupation = replace(occupation, occupation == '?', NA),
         native_country = replace(native_country, native_country == '?', NA))


# Count NAs by variable
na_by_column(df_replaced)

```
**NA_IF**
But there is another way, much easier, by the way, using the function `na_if()`. Simply put, if a value is a given pattern, make it `NA`.
```{r}

# Replacing values "?" with NA
df_replaced <- df %>% na_if('?') 


# Count NAs by variable
na_by_column(df_replaced)


```



#### Filling
Fill values is when you have a `NA` value that needs to be filled with an average or another value, like zero.

```{r}

# Fill NA values with last or next valid value
df_replaced %>% fill(workclass, occupation, native_country,
                     .direction= 'down' )

# Finding the most frequent entry
m_freq_workcls <- names(table(df$workclass)[which.max(table(df$workclass))])
m_freq_occup <- names(table(df$occupation)[which.max(table(df$occupation))])

# Replace NA with the most frequent value
df_no_na <- df_replaced %>%
  replace_na(list(workclass= m_freq_workcls,
                  occupation= m_freq_occup) )


# Drop NAs and save the dataset in a new variable name
df_no_na <- df_no_na %>% 
  drop_na()


```

**Filling NAs from Numeric columns**

```{r}

# Data frame
df_num <- data.frame(A= c(1,2,2,2,3,NA),
                     B= c(3,4,5,3,NA,0),
                     C= c(1,1,1,NA,NA,5))

# Fill NAs for numeric columns with statistics measurements
df_num %>% replace_na(list(A= median(df_num$A, na.rm = T),
                           B= mean(df_num$B, na.rm = T),
                           C= min(df_num$C, na.rm = T) ) )

```


### Arranging Data
Arrange data is useful to rank items from top to bottom or vice-versa, creating ordinated datasets.

```{r}

# Arrange data in increasing order
df_no_na %>% arrange(native_country)

# Arrange data in decreasing order
df_no_na %>% arrange( desc(native_country) )


```

Using `tidyverse` to group variables and order them.

```{r}

# Group and order average net gain by education level
df_no_na %>%
  group_by(education) %>% 
  summarise(count=n(),
            avg_net_gain= mean(capital_gain - capital_loss)) %>% 
  arrange( desc(avg_net_gain) )

```


### Creating New Variables
During data wrangling, it is common to create new variables, being that for splitting an information or combining two or more variables to make a new one.

*Note: to create new columns in R, you can use `df$new_col_name` or `df['new_col_name]`*

```{r}

# Split variable target into sign and amount
df_no_na %>% separate(target, into=c('sign', 'amt'), sep='\\b')

# Unite variables sex, race and age
df_no_na %>% 
  unite(sex, race, age, col='description', sep='_', remove=FALSE)

```

**Creating a custom calculation**
Imagine that there were a rule in place that, for every `capital_gain` - `capital_loss` equal or greater than 15,000 there was a 10% tax.
We will first create a `total` variable to show arithmetic operation and then calculate the 10% tax

```{r}

# Tax variable creation
df_no_na %>% 
  mutate(total_gain = capital_gain - capital_loss,
         tax= ifelse(total_gain >= 15000, 
                     total_gain *0.1,
                     0)  ) %>% 
  arrange(desc(tax))


```

Let’s change <=50K to *under* and >50K to *over*.

```{r}

# Change values from target to over or under
df_no_na %>% 
  mutate( over_under= 
            recode(target, '<=50K'='under', '>50K'='over')    ) %>% 
  select(target, over_under)

```

Mutate and cut to see which observations are over or unde the average.

```{r}

# Observations over or under the average
df_no_na %>% 
  mutate( age_avg = mean(age),
          over_under_age_avg= cut(age,
                                  c(0, mean(age), max(age)),
                                  c('Lower than avg', 'Above the avg'))  ) %>% 
  select(age, age_avg, over_under_age_avg)

```

### Reshape
Reshaping means to transform the dataset from wide format to long or vice versa.
Roughly, we can say that:
Wide format: variables are rows.
Long format: variables are columns.

```{r}
# set seed to generate the same random numbers every run
set.seed(42)

# Create a dataset
df_wide <- data.frame(
  project = c('project1', 'project2','project3'),
  Jan= sample(1000:2000, 3),
  Feb= sample(1000:2000, 3),
  Mar= sample(1000:2000, 3)
)
  
# Wide to Long
df_long <- df_wide %>% 
  pivot_longer(cols= 2:4,
               names_to = 'months',
               values_to = 'expenses')

# View in RStudio viewer
View(df_long)

```

**Long to Wide**
```{r}

# Long to Wide
View(df_long %>%
  pivot_wider(names_from = 'months',
              values_from = 'expenses'))

```







