---
title: "Chapter8_do_more_tidyverse"
author: "Gustavo R Santos"
date: '2022-07-28'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 8 - Transformations with `tidyverse`

This document is part of the Packt Book *Data Wrangling with R*.

---

### Do more with **tidyverse**

In this file, we will go over some interesting functions of the package `tidyverse`.

```{r}
library(tidyverse)
```


### Load a dataset
```{r}

data("mtcars")

```



### Functions from purrr
Functions similar to apply family

**Mapping functions**

`map()`: applies function to every element of list or vector. Returns same data type.


```{r}

# Map
mtcars %>% 
  select(hp, wt) %>% 
  map(mean)

# Map int
mtcars %>% 
  select(hp, wt) %>% 
  map_int(length)



```

### Binding Data

Binding data is useful to glue pieces of data together.

`brind rows` glues the data by columns. So if your data has the same variables, you can use this function.

```{r}

# Creating datasets A and B
A <- mtcars[1:3, ]
B <- mtcars[4:6, ]

# Bind rows
AB <- A %>% bind_rows(B)
AB

```

`bind_cols()` is to bind columns, gathering the data by rows. It is more suitable when the rows are for the same observation, but the columns are different and need to be put together.

```{r}

# Creating datasets A and B
A <- mtcars[1:5, 1:3]
B <- mtcars[1:5, 4:6]

# Bind columns
AB <- A %>% bind_cols(B)
AB
```

**Cumulative functions**

Set of functions used along with `mutate()`, like `cumsum()`, `cumprod()`, `cummean()` and `cummax()`. These are used for cumulative calculations of a variable using a simple syntax, requiring that we just pass a variable name to it within the mutate function, as seen next.

```{r}

# Cumulative sum of weight
mtcars %>% mutate(
  cumulative_sum = cume_dist(wt)
) %>% arrange(cumulative_sum)

```



**Case When**
Function `case_when()` to deal with multiple cases of logical tests.

```{r}

# Case When to create a new label column for the transmission variable
mtcars %>% 
  mutate( transmission_type= 
            case_when(
              am == 0 ~ 'automatic',
              am == 1 ~ 'manual'  )  ) %>% 
  select(am, transmission_type)
```



**ggplot2**

A basic plot with `ggplot2`
```{r}

# ggplot2 basic scatterplot
ggplot(data= mtcars) +
  geom_point( aes(x=hp, y= mpg),  
              color='royalblue', size=4, alpha=0.5 ) +
  ggtitle('Relationship between HP vs. MPG')

```



