---
title: "Chapter 3 - Basic Data Visualization"
author: "Gustavo R Santos"
date: '2022-06-29'
output: html_document
---

#Packt Book
##Data Wrangling With R
###Chapter 2 - How to Load Files to RStudio

This document is part of the Packt Book *Data Wrangling with R*.

---

*Library Import and dataset load*
```{r}

library(datasets)
library(tidyverse)

# Dataset Load
data("mtcars")

```

1. Univariate Visuals

**Histograms**
```{r}

# Histogram of miles per gallon
hist(mtcars$mpg)

```


Customized histogram

```{r}

#Custom histogram
hist(mtcars$mpg, col= 'royalblue',
     main='Histogram of MPG', xlab = 'MPG',
     breaks= 'FD')

```


**Boxplots**

```{r}

# Plot Boxplot of MPG
boxplot(mtcars$mpg, col='royalblue',
        main= 'Boxplot of MPG',
        ylab= 'Miles per Gallon')

```


*Do cars with automatic transmission consume make less miles per galon than the manual vehicles?*
```{r}

# boxplot AM vs MPG
boxplot(mtcars$mpg[mtcars$am == 0],
        mtcars$mpg[mtcars$am == 1],
        col= c('coral', 'royalblue'),
        names= c('Automatic', 'Manual'),
        ylab= 'MPG',
        main= 'Types of Transmission and the Fuel Efficiency'
        )

```

Boxplot using formula notation

```{r}

# Boxplot with formula notation
boxplot(mpg ~ cyl, data= mtcars,
        col= c('gold', 'coral', 'blue'),
        main= 'MPG by number of Cylinders')

```


**Density Plot**
```{r}

# Density calculation
dp <- density(mtcars$mpg)

# Plot
plot(dp, main= 'Density Plot of MPG')

# Fill shape
polygon(dp, col="red")

```


Compare two density plots
```{r}

# Density calculation
dp0 <- density(mtcars$mpg[mtcars$am == 0])
dp1 <- density(mtcars$mpg[mtcars$am == 1])

# Plot
plot(dp0, main= 'Density Plot of MPG')
plot(dp1, main= 'Density Plot of MPG', ylim= c(0,0.1) )
# Legend
legend(x=35, y=0.1 , c('Automatic', 'Manual'), fill= c('red', 'blue'))

# Fill shape
polygon(dp0, border='red', lwd=2)
polygon(dp1, border='blue', lwd=2)

```

**2. Two Variable Plots**

**Scatterplots**

```{r}

# Basic scatterplot
plot(x=mtcars$hp, y=mtcars$mpg, 
     main= 'Scatterplot HP vs. MPG',
     xlab= 'HP', ylab= 'MPG',
     col='royalblue', pch=16)

```


```{r}

# MPG vs Weight
plot(x=mtcars$wt, y=mtcars$mpg, 
     main= 'Scatterplot Weight vs. MPG', cex= 2,
     xlab= 'HP', ylab= 'MPG',
     col='royalblue', pch=16)

```


**Bar/ Column Plot**

```{r}
# Column Plot: How many cars by cylinder
cyl_counts <- table(mtcars$cyl)
barplot(cyl_counts, col='royalblue',
        main='Number of Cars by Cylinder')

```


```{r}
# Bar Plot
v_eng <- table(mtcars$vs)
barplot(v_eng, main='Cars with V-Engine vs Straight',
        horiz=TRUE, col='royalblue',
        names= c('V-Shaped', 'Straight'))


```


**Line Plot**

```{r}

# Load dataset
data("AirPassengers")

# Plot line
plot(AirPassengers, col= 'skyblue',
     lwd=3, main= 'Evolution of the Number of Air Passengers from 1949-1960')

```

```{r}

# Random X and Y
x <- 1:12
y <- rnorm(12)+2
y2 <- rnorm(12)+3
y3 <- rnorm(12)+4
# Line Plot
plot(x,y, col='royalblue', type= 'l', lwd=3, ylim=c(0,6))
# Add second line
lines(x,y2, col='coral', type= 'b', lwd=3)
# Add third line
lines(x,y3, col='darkgreen', type= 'l', lwd=3, lty = 2)
#Add legend
legend(x=11, y=6, c('y', 'y2', 'y3'), fill=c('royalblue','coral', 'darkgreen'))

```

**3. Plotting Multiple Variables**

We can use more than two variables to plot graphics when we use size or color based on a third variable to differentiate data points.

```{r}

# Scatterplot 3 variables
plot(mtcars$hp, mtcars$mpg, 
     col=mtcars$cyl, cex=mtcars$cyl/3, lwd=2,
     xlab='HP', ylab='MPG',
     main= 'MPG versus HP and Cylinder')
#Add legend
legend(260, 32, sort(unique(mtcars$cyl)), 
       col=sort(unique(mtcars$cyl)),
       cex = sort(unique(mtcars$cyl)/6), 
       pch=1, horiz = T, yjust = 0.5)

```

**Bar Plot with three variables**
Too much information.

```{r}

# Stacked Bar Plot with Colors and Legend
counts <- table(mtcars$vs, mtcars$cyl)
barplot(counts,  xlab="Cylinders", col=c("royalblue","coral"),
        legend.text = c('V', 'Line'),
        args.legend = list(x='top', horiz=T),
        main='Cars by Cylinder and Engine/ Transmission',
        sub= list('Engines with 8 cylinders tend to be constructed in V shape with automatic transmission', col='darkgray'))
lines(factor(c(4,6,8)), table(mtcars$am, mtcars$cyl)[1,], type='b', col='green', lw=3)
lines(factor(c(4,6,8)), table(mtcars$am, mtcars$cyl)[2,], type='b', col='yellow', lw=3)
legend(1.3,12, horiz = T, c('Auto', 'Manual'), fill=c('green', 'yellow'))



```




