---
title: "Chapter6_Practice"
author: "Gustavo R Santos"
date: '2022-07-15'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 6 - Working with Date and Time objects

This document is part of the Packt Book *Data Wrangling with R*.

---

### Import libraries and Load Dataset
```{r}

library(tidyverse)
library(lubridate)

```

## Practice


The dataset to be used in this exercise is the Classic Rock, from fivethirtyeight.
To access the original dataset, got to: https://github.com/fivethirtyeight/data/tree/master/classic-rock
```{r}

# URL where the data is stored
url <- 'https://raw.githubusercontent.com/fivethirtyeight/data/master/classic-rock/classic-rock-raw-data.csv'

# Load to RStudio
df <- read_csv(url)

# Make a copy of the original data
df_original <- df

# View data frame in RStudio viewer
#View(df)
  
```

Transform column time in datetime.
You may have just seen that the column time brings integer numbers.
To be able to work with these dates, we must, first, transform that variable to a datetime variable.

```{r}

# Variable TIME to datetime
df$TIME <- as_datetime(df$TIME)

```

For our purpose, there is no need of all the variables. Let's select some.
```{r}

# Select variables
df <- df[,c('Song Clean', 'ARTIST CLEAN', 
            'CALLSIGN', 'TIME', 'COMBINED', 
            'First?')]

```

Next, we are creating some extra granularity for time, by adding columns for month, day, week day and hour for further analysis.
Notice that year is always 2014 and month is 06, so no point in keeping them.

```{r}

# Add new column year
df$year <- year(df$TIME)
# Add new column month
df$month <- month(df$TIME)
# Add new column day
df$day <- day(df$TIME)
# Add new column week day
df$weekday <- wday(df$TIME)
# Add new column hour
df$hour <- hour(df$TIME)

```


### Visualizations

Since now we have many slices of time, we can create many different views of the musics using that data.

What is the distribution of musics played per day of the week?
```{r}

# Filter only unique observations
df_unique <- df[!duplicated(df$`Song Clean`),]

# Songs by weekday
song_by_wkd <- table(df_unique$weekday)

# Bar plot of Songs by weekday
barplot(song_by_wkd, col='royalblue',
        main='Number of unique songs by day of the week',
        xlab='Day of the Week [1=Sun, 7=Sat]',
        ylab= 'Distinct Songs')

```

What is the hour when there are more music being played?
In this case, I don't need to filter only distinct song, as I really want to know what is the most busy hour of those days.
```{r}

# Songs by hour
song_by_hour <- table(df$hour)

# Bar plot of Songs by hour
barplot(song_by_hour, col='royalblue',
        main='Songs by hour',
        xlab='Hour',
        ylab= 'Songs Played')

```


It looks like there is something weird going on at 11pm, or 23 hours. Let's filter that and have a closer look, just like in a real life project.

```{r}

# Filter and arrange only songs played at 23h
df_23 <- df[df$hour == 23,]
df_23 <- arrange(df_23, `Song Clean`,CALLSIGN, TIME)


```

A first verification can be searching for duplicate entries.
```{r}

# Checking if there are duplicated rows
# Number of rows in the dataset - Number of unique rows
dim(df_23)[1] - df_23 %>% distinct() %>% nrow()

# Filter only the duplicated rows
dups <- df_23[duplicated(df_23),]

# Dimensions of dups
dim(dups)

# Test for real duplicates: if sum of TRUE equal cells is 11 (all columns) then it is dup.
real_dup <- c()
for (row in 1:279) {
  equal_cols = sum(dups[row,] == dups[row+1,])
  if (equal_cols == 12) {real_dup <- c(real_dup, row) }
}

# Look at duplicates
head(dups[c(9,81),])

# Checking duplicates
df_original %>% 
  filter( (TIME == ymd_hms("2014-06-19 23:54:25")) &
          (COMBINED %in% c('Another One Bites the Dust by Queen',
                           'Fly Like an Eagle by Steve Miller Band')) )


# Remove duplicates
df <- df_original[!df_original$UNIQUE_ID %in% c('WCSX0949', 'WCSX1021', 'WCSX0955', 'WCSX1005'),]

# Select variables
df <- df %>% 
  select('Song Clean', 'ARTIST CLEAN', 'CALLSIGN', 'TIME', 'COMBINED', 'First?')
# Variable TIME to datetime
df$TIME <- as_datetime(df$TIME)
# Add new column year
df$year <- year(df$TIME)
# Add new column month
df$month <- month(df$TIME)
# Add new column day
df$day <- day(df$TIME)
# Add new column week day
df$weekday <- wday(df$TIME)
# Add new column hour
df$hour <- hour(df$TIME)

```


Everything looks fine. What happened is that we summed all the songs played by hour for all the days in that week. A better approach should be grouping the dataset by day and take an average of songs played by hour.

```{r}

# Group data by weekday
df_by_day <- df %>% 
  group_by(weekday, hour) %>% 
  summarise(songs_ct = n() )

# Group the result by hour
df_by_day <- df_by_day %>% 
  group_by(hour) %>% 
  summarize(avg_songs_ct = mean(songs_ct)) %>% 
  ungroup()

# Line plot of Songs by weekday
plot(y= df_by_day$avg_songs_ct, df_by_day$hour,
     type = 'l', pch = 19, lwd=3,
     col='royalblue',
     main='Avg Number of Songs Played by hour',
     xlab='Hour',
     ylab= 'Songs Played')

```
Even then, the number of songs played at 11pm is really high.

Now, let's see what hour has most songs being played for the first time.
```{r}

# Filter only first time = 1
first_time <- df[df$`First?`==1,]

# First Time Songs by Hour
first_time <- first_time %>% 
  group_by(hour) %>% 
  summarise(song_ct= n())


# Line plot of First appearance Songs by hour
plot(x=first_time$hour, y=first_time$song_ct,
     col='royalblue', type='l', lwd=3,
        main='First Time Played by hour',
        xlab='Hour',
        ylab= 'Songs Played')


```

Average Number of songs played by radio by day
```{r}

# Average by radio by day
by_radio <- df %>% 
  group_by(CALLSIGN, weekday) %>% 
  summarise(song_ct = n()) %>% 
  group_by(CALLSIGN) %>% 
  summarise(avg_song= mean(song_ct)) %>% 
  arrange(desc(avg_song)) %>% 
  ungroup() %>% 
  head(5)

# Bar Plot
barplot(avg_song ~ CALLSIGN, data=by_radio,
        horiz = T,
        main='Average Number of Songs by Radio Station in a Day',
        xlab = 'Avg number of songs played', ylab='Radio Station',
        col='royalblue')

```






