---
title: "Chapter6_Working_with_DateTime"
author: "Gustavo R Santos"
date: '2022-07-14'
output: html_document
---

## Packt Book
## Data Wrangling With R
### Chapter 6 - Working with Date and Time objects

This document is part of the Packt Book *Data Wrangling with R*.

---

### Import libraries and Load Dataset
```{r}

library(tidyverse)
library(lubridate)

```

## Introduction

There are three main ways to work with datetime objects in data science.
1. You create them
2. You parse them from a string or text
3. Arithmetic operations

---

**1. Creating a Date or datetime objects**
Computers register date and time in their systems. As convention, computers compute any date and time objects using January 1st, 1970 at 00:00:00 UTC.
You can check the current date and time from your system using *lubridate*.
```{r}

# Checking today's system date
today()

# Checking current system's time
now()

# Creating a date object
as_date(0)
ymd(20220714)
mdy('Jul142022')
yq('2010Q4')

# Creating a time object
hms::as_hms(43200)
ymd_hms(20220714150000)

# Creating a datetime object one year after 1970-01-1
#(60s*60min*24h*365d = 31,536,000 seconds)
as_datetime(31536000)

```

**datetime to date**
Dropping time portion of a datetime object.
```{r}

# Datetime object
dt_tm <- ymd_hm( '2022-01-02 03:04' )

# Convert to just date object
as_date( dt_tm)

```


**2. Parsing datetime objects from strings or text**

When dealing with datetime objects in a text or data frame and you want to parse or split the dates, use the following commands.

```{r}

# Creating a variable with 4 dates
dt <- c('2000-01-01 01:01:01',
        '2005-02-02 02:02:02',
        '2010-03-03 03:03:03',
        '2020-04-04 04:04:04')
# Assign it as datetime object
dt <- as_datetime(dt)

### From a variable of dates, split them ###

# years
year(dt)
# months
month(dt)
# weeks
week(dt)
#week day
wday(dt)
# days
day(dt)
# hours
hour(dt)
# minutes
minute(dt)
# seconds
second(dt)
# time zones
tz(dt)

```
**Gather date/ time objects**
```{r}

# Separate variables
y_obj <- '2022'
m_obj <- '5'
d_obj <- '10'

# gather date
ymd( paste(y_obj,m_obj,d_obj, sep='-') )


```


**3. Arithmetic Operations**

Arithmetic operations with datetime are slightly more complex than operations with real numbers. The complexity comes from different units and specific properties, such as time zones.

If we add 20 days to Jan 10, we get to the 30th. But if we add it to Feb, we will land in March, for example.
```{r}

# Simple subtraction
dt1 <- as_date('2022-06-01')
dt2 <- as_date('2022-05-01')

dt1 - dt2

```


To comprehend this, we must understand periods, durations and intervals.
* Period: tracks changes in time, not accounting for any deviations.
* Duration: tracks passage of time, accounting for deviations like gaps.
* Interval: a time interval composed by start and end.

**Period**
```{r}

# Date
dt <- ymd('2000-01-01')

# Date minus a past date
today() - dt

# Create a period to add or subtract
p <- years(x=22) + months(x=06) + days(x=13)
# Another syntax
p <- period( c(22, 6, 13), c('year','month', 'day'))
p
# Operations
dt + p
dt - p

#--Next maintenance--
# Date of the last maintenance
dt <- ymd('2021-01-15')

# Create a period to add or subtract
p <- years(x=1) + months(x=06) + days(x=1)
# Another syntax
p <- period( c(1, 6, 1), c('year','month', 'day'))
# Calculation
writeLines( paste('Next maintenance date is on:', dt + p) )


```

**Duration**
```{r}

# Date
dt <- ymd('2000-01-01')

# Create a duration of 5 years
dw <- dyears(x=5)
dw

# Calculate warranty time 5 years after dt
warranty_end <- dt+dw
writeLines( paste('Warranty ends on:', warranty_end) )

```

**Interval**
```{r}

# Date
dt <- ymd('2022-01-01')

# Interval start
i <- interval(start= "2021-01-01", end= "2022-12-31")
i

# Date within interval
dt %within% i

```
**Time Zone**

```{r}

# System time zone
Sys.timezone()

# Creating a datetime object in another timezone
ymd_hms('2022-01-01 00:00:00')
ymd_hms('2022-01-01 00:00:00', tz='Europe/Paris')

#--Display datetime in different time zone--

# Date creation
dt_dubai <- ymd_hms('2022-07-01 10:00:00', tz='Asia/Dubai')
with_tz(dt_dubai, tzone='America/New_York')


```

## Datetime in text

Parsing datetime objects from text.
```{r}

# Lubridate parsing
mdy('The championship starts on 10/11-2000')

```
```{r}

#--Parsing with Regex--

# Text
t <- "The movie was launched on 10/10/1980. It was a great hype at that time, being the most watched movie on the weeks of 10/10/1980, 10/17/1980, 10/24/1980. Around ten years later, it was chosen as the best picture of the decade. The cast received the prize on 09/20/1990."

# Parse using regex
str_extract_all(t, '[0-9]+/[0-9]+/[0-9]+')

```
