---
title: "Chapter2_Load_Files_RStudio"
author: "Gustavo R Santos"
date: '2022-06-20'
output: html_document
---

#Packt Book
##Data Wrangling With R
###Chapter 2 - How to Load Files to RStudio

---

This document is part of the Packt Book *Data Wrangling with R*.

1. Let's load a file from the built-in datasets from R

```{r}
# Loading a preinstalled dataset to R
data("Orange")
```

We can perform the same task, but loading it to a different variable name.

```{r}

# Load the dataset to a different variable name
library(datasets)
df <- datasets::Orange

```


2. We can save the file to a CSV now with the following code.
It will be saved to the same folder where your script or RProj file is saved. To know where it is being saved, use the command `getwd()` on your Console screen.

```{r}

# Save a variable to csv
write.csv(df, 'oranges.csv', row.names = FALSE)
write_csv()
```



3. Next we are going to load the same dataset *Orange* from a CSV file in our local machine.

```{r}
# Load dataset from csv using base R
path = 'C:/Users/.../oranges.csv'
df <-  read.csv(path)
```


4. We can also load a csv file using the *readr* library from the *tidyverse* package.

```{r}
# Load dataset from csv using tidyverse
# Load the library
library(tidyverse)

# setup the file path
path = 'C:/Users/.../oranges.csv'

# Load the file
df <-  read_csv(path)
```

We can list the strings that should be considered missing values (NA).
In the example below, we're considering blank values and "No Record".

```{r}
# Listing entries to consider NA

# setup the file path
path2 = 'C:/Users/.../oranges_modified.csv'

df <- read_csv(path2, na = c('','No Record') )
```

Load a file with restrictions of quantity of rows and/ or columns to be loaded.

```{r}
# Load a file with only the original columns and 10 rows
df <- read_csv(path2, col_select = c(1, 2, 3), n_max = 10)
df %>% head(3)
```



5. Tibble vs Dataframe
5.1 Printing

Tibble can print all the variables in the console screen.

```{r}

# Print all the columns using tibble
data("world_bank_pop")
world_bank_pop %>% print(width=Inf)

```

Or you can also use the View function to access the built-in viewer from RStudio.

```{r}
View(world_bank_pop)
```

5.2 Column Names
data.frame won't accept non-standard column names unless it changes it.
Notice that *my column* will become *my.column*

```{r}
data.frame( `my column` = c(1,2,3) )
```
Tibble accepts it and won't change it.

```{r}
tibble(`my column` = c(1,2,3))

```

